---
title: "Exploratory Data Analysis"
output: pdf_document
---

# Introduction
This document presents an exploratory data analysis of the dataset.

## Dependency Setup
```{r Dependency Setup, inlcude = TRUE}

if (!require("here")) install.packages("here")
library(here)

# Set current working directory to file location
setwd(here::here())

#Print current working directory in case not connected to .Rproj file
print(getwd())

# Define the path to the .renv lockfile within the 'renv' folder
renv_lockfile <- file.path(getwd(), "renv.lock")

# Activate and restore packages from renv lock file
if (!require("renv")) install.packages("renv")
renv::restore()

# Use the p_load function to install and load any desired packages that may not 
# have been in the renv.lock file. Note, if the packages were already in the 
# renv.lock file, that is okay. `p_load` will not change version number

if (!require("pacman")) install.packages("pacman")
pacman::p_load(renv, rmarkdown,  knitr, tidyverse, reshape2, haven, xtable,
               gridExtra, broom, car, caret, Metrics, stargazer)

# Update the renv lock file
renv::snapshot()
```

# Data Loading
```{r}

# Load custom functions
#source("custom_functions.R")

# load data
wave1 <- read_dta("~/Desktop/DataSci/DATA2020-Spring2024/DATA2020_Final_Project/data/raw/FF_wave1_2020v2.dta")
wave2 <- read_dta("~/Desktop/DataSci/DATA2020-Spring2024/DATA2020_Final_Project/data/raw/FF_wave2_2020v2.dta")
wave3 <- read_dta("~/Desktop/DataSci/DATA2020-Spring2024/DATA2020_Final_Project/data/raw/FF_wave3_2020v2.dta")
wave4 <- read_dta("~/Desktop/DataSci/DATA2020-Spring2024/DATA2020_Final_Project/data/raw/FF_wave4_2020v2.dta")
wave5 <- read_dta("~/Desktop/DataSci/DATA2020-Spring2024/DATA2020_Final_Project/data/raw/FF_wave5_2020v2.dta")
wave6 <- read_dta("~/Desktop/DataSci/DATA2020-Spring2024/DATA2020_Final_Project/data/raw/FF_wave6_2020v2.dta")
```

# Merge Waves based on Encrypted Family ID
```{r}
# https://ffcws.princeton.edu/sites/g/files/toruqf4356/files/year_3_guide.pdf

data <- merge(wave1, wave2, by = "idnum")
data <- merge(data, wave3, by = "idnum")
data <- merge(data, wave4, by = "idnum")
data <- merge(data, wave5, by = "idnum")
data <- merge(data, wave6, by = "idnum")



#variable_names <- names(data)
#selected_variables <- grep("^r3", variable_names, value = TRUE)
#print(selected_variables)

```

# Separate Variables into categories and filter
```{r}

# Numeric Variables
numeric_vars <- c("cf3age", "cm3age", "f3l1")

# Race variables of Respondent - variables are one hot encoded

race_vars_nonwave_3  <- c("f1h3", "f2g5", "f4h0c")
race_vars_wave_3 <- c("r3f13_0", "r3f13_1", "r3f13_2", "r3f13_3", "r3f13_4", "r3f13_5", "r3f13_9")
hispanic_wave_3 <- c("r3f14")

race_vars <- c(race_vars_nonwave_3, race_vars_wave_3, hispanic_wave_3)

# Binary Variables
binary_vars <- c("cf3marm", "cf3cohm", "m3i0i", "m3i0j")

# Ordinal variables 
# Check on following variables --- ordinal_vars <- c("m3i0m", "f3i0l1", "m3i0l1", "f3i0m")
ordinal_vars <- c()

# Multi-categorical Variables
multi_categorical_vars <- c("f3e5", "f3i2", "m3c41", "f3c41", "m3i0d", "m3i0e", "m3i0f", "m3i0g", "m3i0h", 
                             "f3i0d", "f3i0e", "f3i0f", "f3i0g", "f3i0h")

# Create vector of desired variables - makes it easier to filter
desired_vars <- c(numeric_vars, race_vars, hispanic, binary_vars, ordinal_vars, multi_categorical_vars)

# Filter out only the numeric variables
data <- data %>% 
  select(all_of(desired_vars))

table(data$r3f13_0)



```


```{r, Numeric Vars}
# Display the structure of the dataset

# Filter data to remove rows with negative values for age and income
data <- data %>%
  filter(if_all(all_of(numeric_vars), ~ . >= 0))

# Summary statistics for numeric variables
summary(select(data, all_of(numeric_vars)))

```

```{r}
# New variables for race and Hispanic status
data$Race_Category <- NA  # Initializes with NA
data$Hispanic <- as.character(ifelse(data$r3f14 %in% c(-9, -3), NA, factor(data$r3f14, levels = c(0, 1), labels = c("No", "Yes"))))

# Update race category logic to combine "Black" categories
for (i in 1:nrow(data)) {
  if (data$f1h3[i] %in% c(-9, -3)) {
    idx <- which(data[i, race_vars_wave_3] == 1)
    if (length(idx) > 0) {
      data$Race_Category[i] <- c("Bi/Multiracial", "White", "Black", "American Indian or Alaska Native", "Asian", "Native Hawaiian or Other Pacific Islander", "Other")[idx]
    } else {
      data$Race_Category[i] <- "No Information"
    }
  } else {
    data$Race_Category[i] <- switch(as.character(data$f1h3[i]),
                                    `1` = "White",
                                    `2` = "Black",
                                    `3` = "Asian",
                                    `4` = "American Indian",
                                    `5` = "Other",
                                    `101` = "Hispanic",
                                    "No Information")
  }
}
data$Race_Category[data$Race_Category == "Black or African American"] <- "Black"

# Update Hispanic status based on better handling of source data
data$Hispanic <- ifelse(data$r3f14 == -9 | data$r3f14 == -3, "No Information",
                        ifelse(data$r3f14 == 1, "Yes", "No"))
data$Hispanic <- factor(data$Hispanic, levels = c("No", "Yes", "No Information"))

# Print to verify
table(data$Race_Category)
table(data$Hispanic)

```


```{r}
# First, you'll create the contingency table
data_table <- table(data$Race_Category, data$Hispanic)

# Convert the table to a data frame
data_frame <- as.data.frame.matrix(data_table)

# Add a race category as a new column
data_frame$Race_Category <- rownames(data_frame)

# Filter out the 'No Information' category
filtered_data <- data_frame[data_frame$Race_Category != "No Information", ]

# Melt the filtered data frame to long format for ggplot2
data_long <- melt(filtered_data, id.vars = "Race_Category")

# Rename columns for clarity
names(data_long) <- c("Race", "Hispanic_Status", "Count")

# Create the stacked bar chart
ggplot(data_long, aes(fill = Hispanic_Status, y = Count, x = Race)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Distribution of Race/Ethnicity by Hispanic Identification (Excluding 'No Information')",
       x = "Race Category",
       y = "Count",
       fill = "Hispanic Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for better visibility


# Display the plot
  ggsave(paste0(output_directory, "/race_ethnicity_chart", var, ".png"), plot = p, width = 10, height = 8, dpi = 300)

```

### Remove messy Race/Ethnicity Variables

```{r}

data <- data %>% 
  select(-all_of(race_vars))

```


```{r}
for (var in race_vars_nonwave_3) {
  labels <- attr(data[[var]], "labels")
  if (!is.null(labels)) {
    cat("Variable:", var, "\n")
    print(labels)
    cat("\n")
  }
}



```



--- Information on Variables Used

**Wave 3 of the Dataset: Political Participation and Civic Engagement Variables**

Mother - 
- `m3i0d`: "In past year, have you participated in: labor union or work-related group?"
- `m3i0e`: "In past year, have you participated in: community organization?"
- `m3i0f`: "In past year, have you participated in: group working with children?"
- `m3i0g`: "Have you ever participated in a political demonstration?"
- `m3i0h`: "In past year have you taken part in a political demonstration?"
- `m3i0i`: "Are you registered to vote?"
- `m3i0j`: "Are you eligible to vote?"
- `m3i0k`: "Did you vote in the November 2000 election?"

Father -
- `f3i0d`: "In past year, have you participated in: labor union or work-related group?"
- `f3i0e`: "In past year, have you participated in: community organization?"
- `f3i0f`: "In past year, have you participated in: group working with children?"
- `f3i0g`: "Have you ever participated in a political demonstration?"
- `f3i0h`: "In past year have you taken part in a political demonstration?"
- `f3i0i`: "Are you registered to vote?"
- `f3i0j`: "Are you eligible to vote?"
- `f3i0k`: "Did you vote in the November 2000 election?"

**Potential Demographic Control Variables:**
  - `cf3age`: Age
  - `r3f13_0`: Bi or multiracial
  - `r3f13_1`: White or Caucasian
  - `r3f13_2`: Black or African American
  - `r3f13_3`: Asian
  - `r3f13_4`: Native Hawaiian or Pacific Islander
  - `r3f13_5`: Native American or Alaskan
  - `r3f13_9`: Other races .
  - `r3f14`: Asking if the respondent is of Hispanic or Latino origin or descent
  - `r3f15`: Asking if the respondent is Mexican, Cuban, Puerto Rican, Dominican, or of other specified Hispanic descent
  - `f3e5`: Which best describes current partner's race?
  - `f3e5a`: Is current partner of Hispanic or Latino descent?
  - `f3e5b`: Is he/she Mexican, Puerto Rican, Cuban, or Other Hispanic?

**Potential Socioeconomic Control Variables:**
  - `cf3edu`: Father's Education Level
  - `cm3edu`: Mother's Education Level
  - `f3l1`: Income (total household income before taxes)
  - `f3i2`: What is your current housing situation?
  - `f3c41` - Is father currently working, in school, or unemployed
  - `m3c41` - Is mother currently working, in school, or unemployed
  - `cf3cohm ` - Father living with child's mother at year three
  - `cf3marm` - Is father married to child's mother at year three?
  - `o3p8` - How would you best describe the home or building?

**Variables not in Dataset but in Codebooks**

- `m3i0l1`: "How important is it: to vote in elections?"
- `m3i0m` : "How important is it to volunteer time to community service?"
- `f3i0l1`: "How important is it: to vote in elections?"
- `f3i0m` : "How important is it to volunteer time to community service?"

---

# Data Cleaning and Overview


```{r}
# Transform from wide to long format
long_data <- pivot_longer(data, cols = starts_with("r3f13"), names_to = "Race", values_to = "Value")
long_data <- long_data[long_data$Value == 1, ]

# Safely add Hispanic/Latino information as a new race category
long_data <- long_data %>%
  mutate(Hispanic = ifelse(Race == "r3f13_0" & data$r3f14[seq_along(Race)] == 1, "Hispanic/Latino", "Not Hispanic/Latino")) %>%
  filter(Hispanic == "Hispanic/Latino" | Race != "r3f13_0")

# Update Race information
long_data$Race[long_data$Hispanic == "Hispanic/Latino"] <- "Hispanic/Latino"

# Creating the plot
ggplot(long_data, aes(x = Race, fill = Race)) +
  geom_bar() +
  labs(title = "Race Demographics", x = "Race Category", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")  # Choosing a color palette
```


```{r, }

# Count levels for categorical variables (including binary and ordinal)
sapply(data[, c(binary_vars, ordinal_vars, multi_categorical_vars)], function(x) list(levels = levels(as.factor(x))))

```

# Data Cleaning 
```{r}
# Handling missing values
data <- data %>%
  mutate_at(vars(all_of(c(numeric_vars, binary_vars, ordinal_vars, multi_categorical_vars))),
            ~ifelse(is.na(.), median(., na.rm = TRUE), .))

# Converting binary variables to factor
data[binary_vars] <- lapply(data[binary_vars], factor)

# Converting ordinal variables to ordered factor
data[ordinal_vars] <- lapply(data[ordinal_vars], function(x) factor(x, ordered = TRUE))


```

# Exploratory Data Analysis - Numeric Variables
```{r}

# Set Output directory
output_directory <- file.path(getwd(), "outputs/EDA")

for(var in numeric_vars) {
  # Generating histogram for each variable
  p <- data %>%
    select(all_of(var)) %>%
    ggplot(aes(x = .data[[var]])) +
    geom_histogram(bins = 30, fill = "blue", color = "black") +
    ggtitle(paste("Histogram of", var))  # Adding title

  # Saving the plot
  ggsave(paste0(output_directory, "/histogram_", var, ".png"), plot = p, width = 10, height = 8, dpi = 300)
}

# Loop through each numeric variable to generate and save boxplot
for(var in numeric_vars) {
  # Generating boxplot for each variable
  p <- data %>%
    select(all_of(var)) %>%
    ggplot(aes(x = var, y = .data[[var]])) +
    geom_boxplot() +
    ggtitle(paste("Boxplot of", var))  # Adding title

  # Saving the plot
  ggsave(paste0(output_directory, "/boxplot_", var, ".png"), plot = p, width = 10, height = 8, dpi = 300)
}

```

# Exploratory Data Analysis - Categorical Variables

```{r}
# Bar plots for categorical variables
p <- data %>%
  select(all_of(c(binary_vars, ordinal_vars, multi_categorical_vars))) %>%
  gather() %>%
  ggplot(aes(x = value)) +
  geom_bar() +
  facet_wrap(~key, scales = "free")

  # Saving the plot
  ggsave(paste0(output_directory, "/barplots_categorical_vars", ".png"), plot = p, width = 10, height = 8, dpi = 300)

```

```{r}
# Loop through each variable to create and save a separate plot
for(var in c(binary_vars, ordinal_vars, multi_categorical_vars)) {
  p <- data %>%
    select(all_of(var)) %>%
    gather() %>%
    ggplot(aes(x = value)) +
    geom_bar() +
    labs(title = paste("Bar Plot of", var), x = var, y = "Count") +
    theme_minimal()
  
  # Saving each plot with a unique filename based on the variable name
  ggsave(paste0(output_directory, "/barplot_", var, ".png"), plot = p, width = 10, height = 8, dpi = 300)
}
```

# Data Visualization
```{r}
# Code for data visualization

# Correlation matrix plot
data %>% 
  select(all_of(numeric_vars)) %>%
  ggpairs()

# Use chi-squared tests or similar for categorical variables if needed

# Example of a chi-square test for two categorical variables
table <- table(data$var8, data$var9)  # Replace var8 and var9 with your variable names
chisq.test(table)

```
